Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2021-04-07 paper/ifo-introduction.tex
the low-hanging fruit of edits

% CW 
To date, the network of gravitational-wave observatories has observed short-duration transient signals from the mergers of binary black holes~\cite{GW150914,GW151226,GW170104,GW170814,GW190521,GWTC-1:2018,GWTC-2:2020} and binary neutron stars~\cite{GW170817,GW170817multi,GW190425,GWTC-1:2018,GWTC-2:2020}, as well as a black hole merger with an object that could either have been a neutron star or another black hole~\cite{GW190814}. 
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals, which are yet to be detected. 
Continuous waves may be emitted at a constant frequency or may wander slowly in frequency over time. 
Rotating neutron stars are prime candidates for continuous-wave emission, especially those in low mass X-ray binaries (LMXB), where the neutron star is in orbit with a low mass stellar companion.
%where a compact object such as a black hole or a neutron star is in orbit with a low mass stellar companion.
% we're talking about neutron stars in LMXB's, so we don't need to mention black holes in LMXB's.
The rotation frequency of the neutron star can wander over time due to variable accretion of matter (and hence angular momentum transfer) from the stellar companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is a prime LMXB target for continuous wave searches. 
%Scorpius X-1 is an LMXB which is particularly bright in X-rays, making it a prime target for continuous wave searches. %from LMXBs. 
Numerous, as yet unsuccessful, searches have been performed for Scorpius X-1 and other LMXBs~\cite{ScoX1O2Viterbi:2019, ScoX1ViterbiO1:2017, RadiometerO1O2:2019, SearchRadiometerO1:2017, SearchCrossCorrO1:2017, ScoX1ViterbiO1:2017, SearchTwoSpecS6:2017, MiddletonEtAlO2LMXBs:2020}.

% changed present to use, as this is not the first time a Michelson has been presented as a tpy GW detector - hannah
In this paper, we use a table-top Michelson interferometer as a toy gravitational-wave detector designed to detect sound instead of gravitational waves. 
We then extend its use to an ``optical microphone'' (using light to capture sound) and present a range of example analysis techniques for educators to use. 
% re-added "as an undergraduate experiment to be clear we are not talking about school teaching - hannah
As an undergraduate lab experiment, the demonstration can be used to teach topics ranging from continuous-wave detection and analysis to electronics, signal processing, and speech enhancement. 
%% I don't like the use of ``freedom to explore''. 
% removed Additionally - I don't think it is neccessary - hannah
It allows students in courses such as physics and electrical engineering to explore the response of an accessible, yet nontrivial, optomechanical system using a hierarchy of data analysis techniques of increasing complexity, including those used in the search for continuous waves in LIGO and Virgo data~\cite{ScoX1O2Viterbi:2019, SuvorovaEtAl:2017,SuvorovaEtAl:2016}. 
% This demonstration can be adapted to an effective outreach tool to explain gravitational wave science to a non-specialist audience.
% Rephrased as this kind of sounds like we are an exhibition too [hannah] - This demonstration also has the potential to be used as an outreach tool alongside other exhibitions
%This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities including; hands-on demonstrations and exhibitions~\cite{TTExhibit:2020,CavagliaExhibit:2009,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, phone applications~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online} which have been developed by gravitational wave research groups around the world. 
This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities developed by gravitational-wave research groups around the world~\cite{TTExhibit:2020,CavagliaExhibit:2009, LIGOScienceEducationCentre:online, GravityDiscoveryCentre:online, GWOSC:online, LOSC:2015, LaserLabs:online, SciVR:online,chirp:online, BlackHoleHunter:online, ArthurJeffesMusic:online, GravitySynthLeonTrimble:online}. 
These tools allow scientists to cater to the increased public and media interest in this field and explain gravitational-wave science to non-specialist audiences.
%\han{look up Sarah Farmer's project}

This paper is laid out as follows. 
In Section~\ref{sec:ifo}, we detail the table-top interferometer design. 
In Section~\ref{sec:single_tone}, we demonstrate observing a single tone from a speaker. 
In Section~\ref{sec:viterbi_wandering}, we replace the single tone with a wandering frequency signal and analyse the signal using a hidden Markov model technique (the Viterbi algorithm) from continuous-wave searches. 
In Section~\ref{sec:optical_microphone}, we demonstrate capture and playback of complex audio, such as music and speech.
% removed as there's not space to really say what this means here: to varying degrees of success.
This demonstration of an optical microphone serves as a more general exhibition of signal processing with a range of examples which can be used in the undergraduate laboratory. 
We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}. 
We collect the software and scripts used to produce this work in Appendix~\ref{app:code}.

>>>>

% CW 
To date, the network of gravitational-wave observatories has observed short-duration transient signals from the mergers of binary black holes~\cite{GW150914,GW151226,GW170104,GW170814,GW190521,GWTC-1:2018,GWTC-2:2020} and binary neutron stars~\cite{GW170817,GW170817multi,GW190425,GWTC-1:2018,GWTC-2:2020}, as well as a black hole merger with an object that could either have been a neutron star or another black hole~\cite{GW190814}. 
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals, which are yet to be detected. 
Continuous waves may be emitted at a constant frequency or may wander slowly in frequency over time. 
Rotating neutron stars are prime candidates for continuous-wave emission, especially those in low mass X-ray binaries (LMXB), where the neutron star is in orbit with a low mass stellar companion.
%where a compact object such as a black hole or a neutron star is in orbit with a low mass stellar companion.
% we're talking about neutron stars in LMXB's, so we don't need to mention black holes in LMXB's.
The rotation frequency of the neutron star can wander over time due to variable accretion of matter (and hence angular momentum transfer) from the stellar companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is a prime LMXB target for continuous-wave searches. 
%Scorpius X-1 is an LMXB that is particularly bright in X-rays, making it a prime target for continuous wave searches. %from LMXBs. 
Numerous, as yet unsuccessful, searches have been performed for Scorpius X-1 and other LMXBs~\cite{ScoX1O2Viterbi:2019, ScoX1ViterbiO1:2017, RadiometerO1O2:2019, SearchRadiometerO1:2017, SearchCrossCorrO1:2017, ScoX1ViterbiO1:2017, SearchTwoSpecS6:2017, MiddletonEtAlO2LMXBs:2020}.

% changed present to use, as this is not the first time a Michelson has been presented as a tpy GW detector - hannah
In this paper, we use a table-top Michelson interferometer as a toy gravitational-wave detector designed to detect sound instead of gravitational waves. 
We then extend its use to an ``optical microphone'' (using light to capture sound) and present a range of example analysis techniques for educators to use. 
% re-added "as an undergraduate experiment to be clear we are not talking about school teaching - hannah
As an undergraduate lab experiment, the demonstration can be used to teach topics ranging from continuous-wave detection and analysis to electronics, signal processing, and speech enhancement. 
%% I don't like the use of ``freedom to explore''. 
% removed Additionally - I don't think it is neccessary - hannah
It allows students in courses such as physics and electrical engineering to explore the response of an accessible, yet nontrivial, optomechanical system using a hierarchy of data analysis techniques of increasing complexity, including those used in the search for continuous waves in LIGO and Virgo data~\cite{ScoX1O2Viterbi:2019, SuvorovaEtAl:2017,SuvorovaEtAl:2016}. 
% This demonstration can be adapted to an effective outreach tool to explain gravitational wave science to a non-specialist audience.
% Rephrased as this kind of sounds like we are an exhibition too [hannah] - This demonstration also has the potential to be used as an outreach tool alongside other exhibitions
%This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities including; hands-on demonstrations and exhibitions~\cite{TTExhibit:2020,CavagliaExhibit:2009,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, phone applications~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online} which have been developed by gravitational-wave research groups around the world. 
This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities developed by gravitational-wave research groups around the world~\cite{TTExhibit:2020,CavagliaExhibit:2009, LIGOScienceEducationCentre:online, GravityDiscoveryCentre:online, GWOSC:online, LOSC:2015, LaserLabs:online, SciVR:online,chirp:online, BlackHoleHunter:online, ArthurJeffesMusic:online, GravitySynthLeonTrimble:online}. 
These tools allow scientists to cater to the increased public and media interest in this field and explain gravitational-wave science to non-specialist audiences.
%\han{look up Sarah Farmer's project}

This paper is laid out as follows. 
In Section~\ref{sec:ifo}, we detail the table-top interferometer design. 
In Section~\ref{sec:single_tone}, we demonstrate observing a single tone from a speaker. 
In Section~\ref{sec:viterbi_wandering}, we replace the single tone with a wandering frequency signal and analyze the signal using a hidden Markov model technique (the Viterbi algorithm) from continuous-wave searches. 
In Section~\ref{sec:optical_microphone}, we demonstrate capture and playback of complex audio, such as music and speech.
% removed as there's not space to really say what this means here: to varying degrees of success.
This demonstration of an optical microphone serves as a more general exhibition of signal processing with a range of examples that can be used in the undergraduate laboratory. 
We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}. 
We collect the software and scripts used to produce this work in Appendix~\ref{app:code}.

### 2021-04-07 paper/ifo-singleTone.tex
moving footnotes to endnote, starting on changing US spelling
the low-hanging fruit of edits

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{\label{fig:webcam_spectrum}
Recovery of a tone at a constant frequency: Here we plot the Fourier amplitude of the intensity pattern against frequency.
The injected signal has a frequency of $2.09\,{\rm Hz}$, while the recovered signal peaks at $2.099\,{\rm Hz}$ and has a FWHM of $0.033\,{\rm Hz}$.
The plot also shows two harmonics at $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ with amplitudes $10.3 \%$ and $9.8 \%$ of the amplitude of the primary peak, respectively.
}	
\end{figure}

% rearranged as I found this section to switch between method and results too much, I've tried this order now: intro, webcam and data specifics, injection and results.

Continuous-wave searches look for nearly monochromatic signals. In this section, we consider a simple sinusoidal tone at a single, constant frequency. %should be single constant frequency *signal*?

As described in Section~\ref{sec:ifo}, the audio signal is played through a speaker fixed to the back of M2. 
The intensity of the interference pattern is measured at a single point on the screen, indicated by the pink star in Fig~\ref{fig:interference_pattern}. 
The webcam records video in three colour channels; red, blue, and green. 
We use the green channel as an approximation of the total intensity produced by the monochromatic green laser.
The webcam samples at a rate of $30\, {\rm Hz}$, which limits the spectral content of observable signals to less than $15\,{\rm Hz}$, the Nyquist frequency.

@@

A tone with a frequency of $2.09\,{\rm Hz}$ is played through the speaker for one minute. 
The amplitude of the Fourier transform of the recovered signal is shown in Fig.~\ref{fig:webcam_spectrum}. 
We measured a peak amplitude at $2.099\,{\rm Hz}$ with a full width half maximum (FWHM) of $0.033\,{\rm Hz}$. 
Two harmonics can also be seen at integer multiples of the peak frequency. 
The peaks at $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ have amplitudes of $0.103$ and $0.098$ as a fraction of the height of the main peak, respectively. They are likely due to the nonlinear response of the system discussed in Section~\ref{sec:ifo} and Appendix~\ref{app:intensity_derivation}.

>>>>

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{\label{fig:webcam_spectrum}
Recovery of a tone at a constant frequency: Here we plot the Fourier amplitude of the intensity pattern against frequency.
The injected signal has a frequency of $2.09\,{\rm Hz}$, while the recovered signal peaks at $2.099\,{\rm Hz}$ and has an FWHM of $0.033\,{\rm Hz}$.
The plot also shows two harmonics at $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ with amplitudes $10.3 \%$ and $9.8 \%$ of the amplitude of the primary peak, respectively.
}	
\end{figure}

% rearranged as I found this section to switch between method and results too much, I've tried this order now: intro, webcam and data specifics, injection, and results.

Continuous-wave searches look for nearly monochromatic signals. In this section, we consider a simple sinusoidal tone at a single, constant frequency. %should be a single constant frequency *signal*?

As described in Section~\ref{sec:ifo}, the audio signal is played through a speaker fixed to the back of M2. 
The intensity of the interference pattern is measured at a single point on the screen, indicated by the pink star in Fig~\ref{fig:interference_pattern}. 
The webcam records video in three color channels; red, blue, and green. 
We use the green channel as an approximation of the total intensity produced by the monochromatic green laser.
The webcam samples at a rate of $30\, {\rm Hz}$, which limits the spectral content of observable signals to less than $15\,{\rm Hz}$, the Nyquist frequency.

@@

A tone with a frequency of $2.09\,{\rm Hz}$ is played through the speaker for one minute. 
The amplitude of the Fourier transform of the recovered signal is shown in Fig.~\ref{fig:webcam_spectrum}. The Discrete Fourier transform was calculated using the NumPy package for Python~\cite{numpy}~\footnote{\url{https://numpy.org/doc/stable/reference/routines.fft.html}}.
We measured a peak amplitude at $2.099\,{\rm Hz}$ with a full width half maximum (FWHM) of $0.033\,{\rm Hz}$.
Two harmonics can also be seen at integer multiples of the peak frequency. 
The peaks at $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ have amplitudes of $0.103$ and $0.098$ as a fraction of the height of the main peak, respectively. They are likely due to the nonlinear response of the system discussed in Section~\ref{sec:ifo} and Appendix~\ref{app:intensity_derivation}. Also note that the noise appears to rise at lower frequencies.

### 2021-04-07 paper/ifo-supplementary-material.tex
the low-hanging fruit of edits

In section V of the main article, we use the inerferoeter as an optical microphone. 
We use signal processing techniques to recover audio signals o speech and music. 
In this supplementary material we provide futher details of a selection of analysis applied to the data which may be considered for undergraduate laboratory examples.  
We start with some na{\"i}ve approaches and then move to traditional signal processing filters (band-passing and cascaded notches). 
We finish by combining these with advanced statistical techniques (Wiener filter and the logMMSE estimator). 
As an illustrative example, we test all of these filters on the same $1\,{\rm s}$ long speech recording.

>>>>

\begin{figure*}
	\includegraphics[width=0.59\textwidth]{figures/setup_pic2.pdf}
	\includegraphics[width=0.4\textwidth]{figures/circuit_pic3.pdf}
	\caption{
Left: A photograph of the optical microphone with an inset of the photodiode. The Michelson interferometer is shown on the left and the circuit and Raspberry Pi on the right. In the main image, the photodiode is placed behind a cloth screen as explained in the main article. The inset shows a face-on view of the photodiode with the cloth screen removed. 
Right: A photograph of the photodiode circuit assembled on a breadboard. The leads from the photodiode enter from the left of the picture. The Raspberry Pi is shown at the top of the picture.
}
	\label{fig:photographs_of_optical_microphone_and_circuit}
\end{figure*}

In Section~V of the main article, we use the interferometer as an optical microphone. Photographs of the physical apparatus of the optical microphone are shown in Fig.~\ref{fig:photographs_of_optical_microphone_and_circuit}.
We use signal processing techniques to recover audio signals of speech and music. 
In this supplementary material we provide further details of a selection of analysis applied to the data which may be considered for undergraduate laboratory examples.  
We start with some na{\"i}ve approaches and then move to traditional signal processing filters (band-passing and cascaded notches). 
We finish by combining these with advanced statistical techniques (Wiener filter and the logMMSE estimator). 
As an illustrative example, we test all of these filters on the same $1\,{\rm s}$ long speech recording.

### 2021-04-07 paper/ifo-tableTopGWs.tex
small consistency edits
the low-hanging fruit of edits

Gravitational-wave detectors such as LIGO and Virgo are large, complex experiments. 
However, their design is fundamentally based on the Michelson interferometer, an optical configuration commonly used in undergraduate laboratories. 
In a Michelson interferometer, laser light is split by a beamsplitter into two perpendicular arms (see Fig.~\ref{fig:ifo_schematic_webcam}). 
Mirrors at the end of each arm reflect the two beams back to the beamsplitter where they recombine to produce an interference pattern.
The resulting interference pattern is dependent on the relative distance travelled by the beams. 
Current generation gravitational-wave observatories have kilometre-scale arms (with arm lengths $4\,{\rm km}$ and $3\,{\rm km}$ at LIGO and Virgo respectively). 
They can measure minuscule changes in distance due to gravitational waves, e.g. the first detection of a binary black hole merger produced a strain of $10^{-21}$, corresponding to a change in distance which is a fraction of the width of a proton.

@@

\begin{figure}
 \begin{center}
  \includegraphics[height=0.3\textwidth, angle=-90]{figures/webcam_still0_star.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
The interference pattern produced by the table-top Michelson interferometer.
This image was taken with the webcam off the beam axis.%, which may explain why the (predicted circular) pattern appears elliptical.
The pink star indicates the point where the intensity time series measurements were taken.
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
}
\end{figure}

@@

% removed as almost identical to the next sentence
%In the following section, we start simply by injecting and recovering a audio signal of constant frequency (a single tone).

>>>>

Gravitational-wave detectors such as LIGO and Virgo are large, complex experiments. 
However, their design is fundamentally based on the Michelson interferometer, an optical configuration commonly used in undergraduate laboratories. 
In a Michelson interferometer, laser light is split by a beamsplitter into two perpendicular arms (see Fig.~\ref{fig:ifo_schematic_webcam}). 
Mirrors at the end of each arm reflect the two beams back to the beamsplitter where they recombine to produce an interference pattern.
The resulting interference pattern is dependent on the relative distance traveled by the beams. 
Current generation gravitational-wave observatories have kilometer-scale arms (with arm lengths $4\,{\rm km}$ and $3\,{\rm km}$ at LIGO and Virgo respectively). 
They can measure minuscule changes in distance due to gravitational waves, e.g. the first detection of a binary black hole merger produced a strain of $10^{-21}$, corresponding to a change in distance which is a fraction of the width of a proton.

@@

\begin{figure}
 \begin{center}
  \includegraphics[height=0.3\textwidth, angle=-90]{figures/webcam_still0_star.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
The interference pattern produced by the table-top Michelson interferometer.
This image was taken with the webcam aligned off the beam axis. %, which may explain why the (predicted circular) pattern appears elliptical.
The pink star indicates the point where the intensity timeseries measurements were taken.
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
}
\end{figure}

@@

% removed as almost identical to the next sentence
%In the following section, we start simply by injecting and recovering an audio signal of constant frequency (a single tone).

### 2021-04-07 paper/ifo-wanderingTone.tex
the low-hanging fruit of edits

Unlike the signal considered in Section~\ref{sec:single_tone}, continuous gravitational-waves may not be monochromatic.
A continuous-wave signal may wander slowly (and randomly) in frequency over time, due to stochastic internal processes in the superfluid interior of isolated neutron stars\cite{MelatosDouglassSimula:2015,Jones:2010}, or variable accretion flows from a stellar companion for neutron stars in binaries as in LMXBs~\cite{BildstenTB:1998}. 
The audio analogue is a tone that wanders in frequency: a note that changes in pitch.

A Fourier transform applied to the whole dataset (as in Section~\ref{todo}) is not well suited to the wandering frequency case, as the signal can be spread across multiple frequency bins making it challenging or perhaps impossible to detect. 
%A different means of detection is needed for a signal which has a randomly wandering frequency. 
The analysis techniques used in this section are inspired by those used to search for continuous waves from spinning neutron stars in LIGO and Virgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}. 
Continuous-wave searches are performed on long datasets, months to years in duration. 
The frequency of the signal can wander significantly over the observation period. 
In this context, ``significantly'' means across multiple frequency bins, where the typical width of a frequency bin is the reciprocal of the total observation time~\cite{JKS:1998,ScoX1O2Viterbi:2019}.
One method to search for a wandering signal is to split the timeseries data into several shorter segments which are analysed individually.%, and then recombine them under some continuity assumption to coherently analyse the whole set. 
% I think the following is more accurate. The antenna beam pattern is used to compute a detection statistic. Which detection statistic to use depends on the type of target. -Hannah

A ``detection statistic'' is used to quantify the likelihood of a signal being present in the data at each frequency and timeseries segment. 
In gravitational-wave data analysis, the detection statistic gives the likelihood of a signal given the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun~\cite{JKS:1998}. 
%Most continuous wave searches use detection statistics to quantify the likelihood of a signal being present in the data at each frequency for each timeseries segment. 
%The detection statistic used depends on the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun, and the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
The choice of detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
In this work, we take the discrete Fourier transform of each timeseries segment and use its amplitude at each frequency as a detection statistic.
This maximizes the likelihood of detecting a sinusoidal signal in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}. 
The detection statistic is normalized for convenience and we use it to form a grid in time and frequency (a spectrogram) which displays the frequency content of the signal over time. 
% moved one sentence up 
%We also normalize the detection statistics for convenience.

@@

The Viterbi algorithm provides a computationally efficient method for finding the optimal path. 
At every $t_i$ all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}). 
Here we describe the algorithm in pseudocode while referring to the schematic in Fig.~\ref{fig:viterbi}.%, where the circles indicate the elements in the spectrogram with the lighter colours corresponding to a higher likelihood. 
The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at time $t_1$, each $f_j$ state can originate from three prior states at time $t_0$ (except for $f_0$ and $f_{N_f}$ which only have two). The paths between these states are indicated by the lines in Fig.~\ref{fig:viterbi}. At each $f_j$ state, we select the path with the highest $A[f(t_0),f(t_1)] {\rm Pr}[f(t_0)]$ value as the most probable path. These choices are highlighted using the black lines in Fig.~\ref{fig:viterbi} while the gray lines show the rejected paths. For example, the most probable connection to the element labelled (a) is the one directly behind it (i.e., $f_4$). Therefore, this path is selected as the best path from $t_0$ to $t_1$ for $f_4$. To allow backtracking at the end, the index of the most probable connection to a node along with the value of the best path to that node is stored, for each node.

@@

In the following section, we use the Viterbi algorithm to recover a slowly wandering signal.

@@

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/viterbiDiagram.pdf}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each circle represents an element in the time-frequency grid starting from the instant $t_0$ on the left and ending at $t_{N_t}$ on the right. 
Vertically, the grid starts from frequency $f_0$ at the bottom and ending at $f_{N_f}$ at the top. 
The color of each circle indicates the likelihood that a signal is present in the box, where a lighter color corresponds to a higher likelihood. 
At each time, the paths can move up one $f$ bin, move down one $f$ bin, or stay at the same $f$ bin (shown by lines in the diagram). 
Each circle has three paths leading to it, indicated by the lines. 
The best path to each circle at each $t_i$ is highlighted in black. 
Some routes through the grid are found to be dead ends, with respect to optimally reaching the other side, such as the path ending at $t_1$ marked (a). 
At $t=t_{N_t}$ the algorithm chooses the terminating frequency circle which has the highest value given by Eqn.~\ref{eqn:recursiveViterbi}, marked (b). 
The Viterbi path is the path leading to this circle, highlighted in orange from (b) to the start at (c). 
}
\end{figure}

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/expt_overlay_2_viterbi_test_webcam.pdf}
	\caption{\label{fig:viterbi_overlay}
Recovery of a wandering tone using the Viterbi algorithm. 
The spectrogram shows the observed frequency (Fourier) amplitude in each time-frequency bin. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path, respectively. 
On the left, before $\sim 15\,{\rm s}$, the signal changes frequency too quickly for the Viterbi algorithm to recover, given a parameter for the rate of change of one frequency bin per time bin. 
At $150\,{\rm s}$ the data appears anomalous, which may be due to some background noise. }
\end{figure*}

To simulate a slowly wandering signal we modulate the frequency sinusoidally with a modulation amplitude that decays with time. 
We use the same set-up as shown in Fig.~\ref{fig:ifo_schematic_webcam} and the output interference pattern is recorded via webcam as in Section~\ref{sec:single_tone}. 
In this section we test the Viterbi algorithm’s ability to recover the wandering signal.

>>>>

Unlike the signal considered in Section~\ref{sec:single_tone}, continuous gravitational waves may not be monochromatic.
A continuous-wave signal may wander slowly (and randomly) in frequency over time, due to stochastic internal processes in the superfluid interior of isolated neutron stars\cite{MelatosDouglassSimula:2015,Jones:2010}, or variable accretion flows from a stellar companion for neutron stars in binaries as in LMXBs~\cite{BildstenTB:1998}. 
The audio analog is a tone that wanders in frequency: a note that changes in pitch.

A Fourier transform applied to the whole dataset (as in Section~\ref{sec:single_tone}) is not well suited to the wandering frequency case, as the signal can be spread across multiple frequency bins making it challenging or perhaps impossible to detect. 
%A different means of detection is needed for a signal which has a randomly wandering frequency. 
The analysis techniques used in this section are inspired by those used to search for continuous waves from spinning neutron stars in LIGO and Virgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}. 
Continuous-wave searches are performed on long datasets, months to years in duration. 
The frequency of the signal can wander significantly over the observation period. 
In this context, ``significantly'' means across multiple frequency bins, where the typical width of a frequency bin is the reciprocal of the total observation time~\cite{JKS:1998,ScoX1O2Viterbi:2019}.
One method to search for a wandering signal is to split the timeseries data into several shorter segments which are analyzed individually.%, and then recombine them under some continuity assumption to coherently analyze the whole set. 
% I think the following is more accurate. The antenna beam pattern is used to compute a detection statistic. Which detection statistic to use depends on the type of target. -Hannah

A ``detection statistic'' is used to quantify the likelihood of a signal being present in the data at each frequency and timeseries segment. 
In gravitational-wave data analysis, the detection statistic gives the likelihood of a signal given the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun~\cite{JKS:1998}. 
%Most continuous wave searches use detection statistics to quantify the likelihood of a signal being present in the data at each frequency for each timeseries segment. 
%The detection statistic used depends on the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun, and the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
The choice of detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
In this work, we take the discrete Fourier transform of each timeseries segment and use its amplitude at each frequency as a detection statistic.
This maximizes the likelihood of detecting a sinusoidal signal in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood} -- we assume that that the noise is Gaussian.
The detection statistic is normalized for convenience and we use it to form a grid in time and frequency (a spectrogram) which displays the frequency content of the signal over time. 
% moved one sentence up 
%We also normalize the detection statistics for convenience.

@@

The Viterbi algorithm provides a computationally efficient method for finding the optimal path. 
At every $t_i$ all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}). 
Here we describe the algorithm in pseudocode while referring to the schematic in Fig.~\ref{fig:viterbi}.%, where the circles indicate the elements in the spectrogram with the lighter colors corresponding to a higher likelihood. 
The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at time $t_1$, each $f_j$ state can originate from three prior states at time $t_0$ (except for $f_0$ and $f_{N_f}$ which only have two). The paths between these states are indicated by the lines in Fig.~\ref{fig:viterbi}. At each $f_j$ state, we select the path with the highest $A[f(t_0),f(t_1)] {\rm Pr}[f(t_0)]$ value as the most probable path. These choices are highlighted using the black lines in Fig.~\ref{fig:viterbi} while the gray lines show the rejected paths. For example, the most probable connection to the element labeled (a) is the one directly behind it (i.e., $f_4$). Therefore, this path is selected as the best path from $t_0$ to $t_1$ for $f_4$. To allow backtracking at the end, the index of the most probable connection to a node along with the value of the best path to that node is stored, for each node.

@@

The key to the Viterbi algorithm is that we can use the signal's correlation in time to improve the signal-to-noise ratio of the frequency of the wandering signal at a particular time. Since the noise is assumed to not be correlated in time, if the detection statistic at a given frequency is high at a particular time and at neighboring frequencies at neighboring times then the likelihood that the given frequency at that particular time is from the wandering signal increases. By iterating this process through the grid for a suitable choice of the window of neighboring frequencies and times, the Viterbi algorithm is well-suited to finding wandering signals.
In the following section, we use the Viterbi algorithm to recover a slowly wandering signal.

@@

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/viterbiDiagram.pdf}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each circle represents an element in the time-frequency grid starting from the instant $t_0$ on the left and ending at $t_{N_t}$ on the right. 
Vertically, the grid starts from frequency $f_0$ at the bottom and ending at $f_{N_f}$ at the top. 
The color of each circle indicates the likelihood that a signal is present in the box, where a lighter color corresponds to a higher likelihood. 
At each time, the paths can move up one $f$ bin, move down one $f$ bin, or stay at the same $f$ bin (shown by lines connecting the circles in the diagram). 
Each circle has three paths leading to it, indicated by the lines. 
The best path to each circle at each $t_i$ is highlighted in black. 
Some routes through the grid are found to be dead ends, with respect to optimally reaching the other side, such as the path ending at $t_1$ marked (a). 
At $t=t_{N_t}$ the algorithm chooses the terminating frequency circle which has the highest value given by Eqn.~\ref{eqn:recursiveViterbi}, marked (b). 
The Viterbi path is the path leading to this circle, highlighted in orange from (b) to the start at (c). 
}
\end{figure}

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/expt_overlay_2_viterbi_test_webcam.pdf}
	\caption{\label{fig:viterbi_overlay}
Recovery of a wandering tone using the Viterbi algorithm.
The spectrogram (heatmap) colouring indicates the value of the detection statistic (here, the logarithm of the absolute value of the discrete Fourier transform) in each time-frequency bin -- with brighter colours indicating higher values. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path, respectively. 
On the left, before $\sim 15\,{\rm s}$, the signal changes frequency too quickly for the Viterbi algorithm to recover, given a parameter for the rate of change of one frequency bin per time bin. 
At $150\,{\rm s}$ the data appears anomalous, which may be due to some transient background noise. }
\end{figure*}

To simulate a slowly wandering signal we modulate the frequency sinusoidally with a modulation amplitude that decays with time. 
We use the same set-up as shown in Fig.~\ref{fig:ifo_schematic_webcam} and the output interference pattern is recorded via webcam as in Section~\ref{sec:single_tone}. 
In this section, we test the Viterbi algorithm’s ability to recover the wandering signal. 
Note that we implicitly approximate the noise at the webcam as white (uniform in frequency) in this implementation of the Viterbi algorithm. This approximation ignores the increase in noise at low frequencies (below 1~Hz) shown in Fig.~\ref{fig:webcam_spectrum} but is a sound approximation for the higher-frequency (around 5~Hz) wandering signals that we study.
